Peptide retention time prediction

### 2014-12-16 retention.tex
Added some references, and modified some sentences

Mass spectrometry (MS) based methods are unquestionably the most popular
techniques for studying the protein content of biological samples.
The palette of protemic applications where such methods are used is
getting increasingly diversified, ranging from identifying and
quantifying proteins in biological samples \cite{Geiger2012} to
determining posttranslational modifications \cite{Huttlin2010} and
elucidating protein interaction networks \cite{Gavin2011}.

Although available in a variety of flavours, most of the widely spread
MS-based workflows consists of the same main steps. First, the
proteins in the mixture of interest are cleaved into peptides using a
sequence-specific enzyme such as trypsin. This step is necessary
because the mass spectrometers are more sensitive for peptides, and
because spectra produced by full-length proteins are difficult to
interpret. In the next step, the resulting peptides are separated
according to their hydrophobicity on a liquid chromatography (LC)
column. As they gradually elute from the column, the peptides are
ionized via electrospray ionization and subjected to mass spectrometry
analysis. The result of these measurements consists of a set of
spectra, which are subsequently used to infer valuable information
about the proteins present in the initial sample.

@@

\subsection{Index-based methods}
\hyphenation{lear-ning}
\hyphenation{indivi-dual}

@@

 Method name & Link & References  \\
 \hline
SSRCalc & \url{http://hs2.proteome.ca/SSRCalc/SSRCalcX.html} & \cite{Krokhin2004, Krokhin2006}, \cite{Spicer2007} \\
BioLCCC & \url{http://theorchromo.ru/} & \cite{gorshkov2006} \\
ANN & Link? & \cite{petritis2003 ,petritis2006improved} \\
RTModel, RTPredict &  \url{http://ftp.mi.fu-berlin.de/OpenMS/documentation/html/TOPP_RTPredict.html}, 
 & \cite{rtpredict, rtpredictImproved} \\
&\url{http://ftp.mi.fu-berlin.de/OpenMS/documentation/html/TOPP_RTModel.html}  & \\

@@

A limitation of SSRCacls is that the algorithm was
optimized for predicting retention times for certain chromatography
setups. Thus, SSRCalc's prediction ability deteriorates for
experimental conditions diverging from the setup on which the
algorithm was optimized \cite{Spicer2007}.

@@

One of the first machine learning approaches for retention time
prediction was based on an Artificial Neural Network
(ANN) \cite{petritis2003}. Initially, each peptide was described by 20
features giving the number of each of the 20 amino acid residues
present in the sequence. The network was trained using $\sim$7000
peptides with known retention times and evaluated using a set of
$\sim$5000 peptides originating from another organism. The algorithm
reported good accuracies, especially taking into account that it only
used information about the amino acid composition of the peptides.

@@

A drawback of the method was however the large amount of training
peptides required, which made the algorithm difficult to retrain for
other chromatographic conditions. This was a serious limitation in the
context of RPLC separation, where a broad variety of chromatographic
setups is commonly used across research laboratories. To address this,
Klammer et al. proposed a predictor based on support vector machines
(SVMs) \cite{klammer2007improving}. Their method required fewer
training peptides and could thus be easily adapted to new
chromatographic conditions, but yielded lower prediction
accuracies. Subsequently, a more elaborate SVM-based method using a
customized kernel function was proposed \cite{rtpredict, rtpredictImproved}. This
approach provided reasonable prediction accuracy, while requiring as
little as 50 peptides for training.

@@

In our group we developed a software package for retention time prediction which we named {\sc
Elude} \cite{elude1, elude2}. One of the main novelty
points of {\sc Elude} wass the calculation of peptide features based on
both a general hydrophobicity index, as well as a retention index
trained for each dataset at hand. This ensured that the algorithm
could be applied to different chromatography setups with minimal loss
in performance. For cases when little training data was available,
{\sc ELUDE} implemented an additional workflow based on a library of
pre-trained retention models. This workflow required only a few tens
of training peptides to select and calibrate the most suitable model
from the library. Both the selection and calibration were performed
using statistical methods robust to outliers.

However, despite all these efforts, it should be pointed out that none
of the retention time prediction methods currently available is able
to provide accuracies comparable to the small shifts observed when
running a sample twice through a chromatography column
(Figure \ref{fig:repr}). Recently, the lack of suitable
representations for secondary structure elements was indicated as one
of the main causes of prediction errors \cite{Reimer2012}.  Most
probably additional factors are involved, and further endeavors are
required to reach the desired level of accuracy.

@@

% This paragraph is consfusing, should be rephrased
The principle of predictability also extends to targeted proteomics,
in SRM analysis the peptides retention time and fragments are
predicted from retention time, often derived from previous shotgun
analysis. As targeted proteomics move into more comprehensive
techniques, such as DIA/SWATH analysis, it becomes increasingly
attractive to use techniques that do not rely on previous observations
of the exact peptide to determine its properties, but instead to use
de-novo predictions from amino acid sequences.

% New technology to learn 
% - ranking-based method
% - deep learning

@@

>>>>

Mass spectrometry (MS) based methods are unquestionably the most
popular techniques for analyzing the protein content of biological
mixtures.  The ever-expanding palette of applications range from
identifying and quantifying proteins in biological samples
\cite{Geiger2012} to determining posttranslational modifications
\cite{Huttlin2010} and elucidating protein interaction networks
\cite{Gavin2011}.

Although available in a variety of flavors, most of the widely spread
MS-based workflows consists of the same main steps. First, the
proteins in the mixture of interest are cleaved into peptides using a
sequence-specific enzyme such as trypsin. This step is necessary
because the mass spectrometers are more sensitive for peptides, and
because spectra produced by full-length proteins are difficult to
interpret. In the next step, the resulting peptides are separated
according to their hydrophobicity on a liquid chromatography (LC)
column. As they gradually elute from the column, the peptides are
ionized via electrospray ionization and subjected to mass spectrometry
analysis. The result of these measurements consists of a set of
spectra, which are subsequently used to infer valuable information
about the proteins present in the initial sample.

@@

\subsection{Index-based methods}
\label{sec:irt}
\hyphenation{lear-ning}
\hyphenation{indivi-dual}

@@

 Method name & Link & References  \\
 \hline
SSRCalc & \url{http://hs2.proteome.ca/SSRCalc/SSRCalcX.html} & \cite{Krokhin2004, Krokhin2006}, \cite{Spicer2007} \\
BioLCCC & \url{http://theorchromo.ru/} & \cite{gorshkov2006} \\
ANN & Link? & \cite{petritis2003,petritis2006improved} \\
RTModel, RTPredict &  \url{http://ftp.mi.fu-berlin.de/OpenMS/documentation/html/TOPP_RTPredict.html}, 
 & \cite{rtpredict, rtpredictImproved} \\
&\url{http://ftp.mi.fu-berlin.de/OpenMS/documentation/html/TOPP_RTModel.html}  & \\

@@

A limitation of SSRCalc is that the algorithm was optimized for
predicting retention times for certain chromatography setups. Thus,
SSRCalc's prediction ability deteriorates for experimental conditions
diverging from the setup on which the algorithm was optimized
\cite{Spicer2007}.

@@

One of the first machine learning approaches for retention time
prediction was based on an Artificial Neural Network
(ANN) \cite{petritis2003}. Initially, each peptide was described by 20
features giving the number of each of the 20 amino acid residues
present in the sequence. The network was trained using $\sim$7000
peptides with known retention times and evaluated using a set of
$\sim$5000 peptides originating from another organism. The algorithm
reported good accuracy, especially taking into account that it only
used information about the amino acid composition of the peptides.

@@

A drawback of the method was however the large amount of training
peptides required, which made the algorithm difficult to retrain for
other chromatographic conditions. This was a serious limitation in the
context of RPLC separation, where a broad variety of chromatographic
setups is commonly used across research laboratories. To address this,
Klammer et al. proposed a predictor based on support vector machines
(SVMs) \cite{klammer2007improving}. Their method required fewer
training peptides and could thus be easily adapted to new
chromatographic conditions, but yielded lower prediction
accuracy. Subsequently, a more elaborate SVM-based method using a
customized kernel function was proposed \cite{rtpredict, rtpredictImproved}. This
approach provided reasonable prediction accuracy, while requiring as
little as 50 peptides for training.

@@

In our group we developed a software package for retention time prediction which we named {\sc
Elude} \cite{elude1, elude2}. One of the main novelty
points of {\sc Elude} was the calculation of peptide features based on
both a general hydrophobicity index, as well as a retention index
trained for each dataset at hand. This ensured that the algorithm
could be applied to different chromatography setups with minimal loss
in performance. For cases when little training data was available,
{\sc ELUDE} implemented an additional workflow based on a library of
pre-trained retention models. This workflow required only a few tens
of training peptides to select and calibrate the most suitable model
from the library. Both the selection and calibration were performed
using statistical methods robust to outliers.

However, despite all these efforts, it should be pointed out that none
of the retention time prediction methods currently available is able
to provide accuracy comparable to the small shifts observed when
running a sample twice through a chromatography column
(Figure \ref{fig:repr}). Recently, the lack of suitable
representations for secondary structure elements was indicated as one
of the main causes of prediction errors \cite{Reimer2012}.  Most
probably additional factors are involved, and further endeavors are
required to reach the desired level of accuracy.

@@

Moreover, targeted proteomics experiments, such as Selected Reaction
Monitoring (SRM) experiments, are dependent on the predictability of a
peptides behavior. In such experiments the retention times and the
suitable transitions are needed as input for the analysis. Frequently,
such information is derived from previous shotgun analysis. As the
capacity of targeted proteomics experiments expands by the
introduction of the more comprehensive data independent acquisition
techniques techniques\cite{Venable2004} (a.k.a. SWATH analysis), such
pre-identification of peptide transitions and retention times
threatens to become a limiting factor. Our current methods to identify
peptides from DIA are only as comprehensive as the previous shotgun
analysis would be. An attractive alternative is to predict retention
time and transition using {\em de novo} predictions from the peptides
amino acid sequences.  Such a system of identifications not dependent
on previous peptide identifications from shotgun proteomics experiment
would lead to a more direct experimental design and hopefully to more
comprehensie coverage. Thus {\em de novo} retention time predictions
have an important role to fill.
%However, to do so we need to make the technique more accurate.

We can see that there are a number of interesting machine learning
techniques that not yet have been used on the problem of predicting
peptides retention time. Here we will bring up one such techniques,
that we feel are particularly good matches, that we have not yet seen
implemented for the problem. Pairwise Machine-learned ranking (MLR)
\cite{liu2009learning} models can be trained to determine which of two
examples that should be better ranked thatn the other. MLR is
traditionally used for construction of ranking models for information
retrieval systems. However, in the case of retention time prediction
we can use such a predictor to predict the elusion order for sets of
peptides, which can be useful in its own \cite{bailey2012instant}, but
can also be converted into absolute retention times, given that the
distribution of their retention times are known.

It is also worth noting that the current trend of processing huge data
repositories, frequently referred to as bigdata, have put quite some
intest in so called deep learning techniques such as belief networks
using restricted multiple layered bolzman machines
(RBM)\cite{salakhutdinov2009deep} or deep neural networks. This
follows the observation that ideathe performance of predictors often
are enhanced if one can structure the prediction into multiple
layers. The drawback of such models are that they require large
training sets. Given the very large sets of peptides with retention
times that are assembled in various repositories, such as peptide
atlas and pride, such methods could probably be implemented for
retention time prediction as well.

% Might be good to add something about that the relatively low
% accuracy in our predictions, there should be some room for
% improvements in the near future.

@@

%%  LocalWords:  hydrophobicity chromatography elute peptides physico
%%  LocalWords:  electrospray spectrometry analytical proteomics RPLC
%%  LocalWords:  posttranslational phosphorylation proteome analytes
%%  LocalWords:  peptide elutes amino tryptic isoelectric SSRCacls
%%  LocalWords:  workflows trypsin hydrophobicities chromatographic
